---
title: "Modelling a problem as SAT"
description: >
  Turning a real question into clauses, using graph colouring: allocating
  variables, encoding constraints, enumerating answers, and proving that no
  answer exists.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modelling a problem as SAT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(zusat)
```

The API of a SAT solver is small and the documentation for it is short. The
hard part is somewhere else: taking a question you actually care about and
turning it into clauses. This article works through one problem end to end —
colouring a graph — because it exercises nearly everything the package does
and because the modelling decisions it forces are the same ones every SAT
encoding forces.

## Variables are just integers

zusat speaks DIMACS, which means a variable is a positive integer and a
literal is that integer, optionally negated. There are no variable names, no
types, no dimensions. If your problem is about vertices and colours, you have
to decide for yourself which integer means "vertex 3 has colour 2".

That decision is most of the work, and the reliable way to make it is to
allocate variables in blocks and write a function that indexes into them:

```{r}
n_vertices <- 10
n_colours <- 3

# variable for "vertex v has colour c"
vc <- function(v, c) (v - 1L) * n_colours + c
```

`vc()` maps the pair `(v, c)` onto `1:(n_vertices * n_colours)` with no gaps
and no collisions. Everything downstream depends on that being right, so it
is worth checking rather than assuming:

```{r}
ids <- as.vector(outer(seq_len(n_vertices), seq_len(n_colours), vc))
c(distinct = length(unique(ids)), expected = n_vertices * n_colours,
  max = max(ids))
```

Getting this wrong is the single most common way to produce a confidently
incorrect answer. An index function that collides silently encodes a
*different problem*, the solver answers that one correctly, and nothing
anywhere reports an error.

## The graph

The Petersen graph: ten vertices, fifteen edges, and a standard example
precisely because it resists the obvious guesses.

```{r}
edges <- rbind(
  # outer pentagon
  c(1, 2), c(2, 3), c(3, 4), c(4, 5), c(5, 1),
  # spokes
  c(1, 6), c(2, 7), c(3, 8), c(4, 9), c(5, 10),
  # inner pentagram
  c(6, 8), c(8, 10), c(10, 7), c(7, 9), c(9, 6)
)
nrow(edges)
```

## Encoding the constraints

A proper colouring is two conditions. Each vertex gets exactly one colour,
and no edge joins two vertices of the same colour.

The first is a cardinality constraint, and `sat_exactly()` encodes it. The
second is a plain clause per edge per colour: not both endpoints can take
colour `c`.

```{r}
colouring_solver <- function(n_colours, edges, n_vertices) {
  vc <- function(v, c) (v - 1L) * n_colours + c
  s <- sat_solver()

  # exactly one colour per vertex
  for (v in seq_len(n_vertices)) {
    sat_exactly(s, vapply(seq_len(n_colours), function(c) vc(v, c), numeric(1)), 1)
  }

  # adjacent vertices differ
  for (i in seq_len(nrow(edges))) {
    u <- edges[i, 1]
    w <- edges[i, 2]
    for (c in seq_len(n_colours)) {
      sat_add(s, c(-vc(u, c), -vc(w, c)))
    }
  }

  s
}

s <- colouring_solver(3, edges, 10)
s
```

The variable count in that print is exactly `10 * 3 = 30`, which is worth
noticing: `sat_exactly()` added no variables of its own. With three literals
per constraint, forbidding every pair is three clauses, so `"auto"` used the
pairwise encoding. Push the numbers up and it switches to a counter that does
introduce auxiliaries — which is where the next section comes in.

## Solving and decoding

```{r}
sol <- sat_solve(s)
sat_status(sol)
```

A solution is a data frame of `variable` and `value`, which is deliberately
not yet an answer to *our* question. Decoding is the mirror of the encoding:

```{r}
decode <- function(solver, n_colours, n_vertices) {
  vc <- function(v, c) (v - 1L) * n_colours + c
  vapply(seq_len(n_vertices), function(v) {
    lits <- vapply(seq_len(n_colours), function(c) vc(v, c), numeric(1))
    which(sat_value(solver, lits))[1]
  }, numeric(1))
}

colouring <- decode(s, 3, 10)
colouring
```

Never trust that without checking it. The encoding could be wrong, the
decoding could be wrong, and both failures look like a plausible answer:

```{r}
all(vapply(seq_len(nrow(edges)), function(i) {
  colouring[edges[i, 1]] != colouring[edges[i, 2]]
}, logical(1)))
```

## Enumerating answers, and why projection matters

There is rarely only one solution, and `sat_solutions()` finds more:

```{r}
s2 <- colouring_solver(3, edges, 10)
ours <- seq_len(10 * 3)

projected <- sat_solutions(s2, vars = ours, limit = 200)
sat_n_solutions(projected)
sat_complete(projected)
```

120 is the right answer: it is the Petersen graph's chromatic polynomial
evaluated at three.

Here `vars` changes nothing, because this encoding introduced no auxiliary
variables to project away. It starts to matter as soon as one does. Compare
the two encodings of the same constraint:

```{r}
pairwise <- sat_solver()
sat_at_most(pairwise, 1:6, 2, encoding = "pairwise")

sequential <- sat_solver()
sat_at_most(sequential, 1:6, 2, encoding = "sequential")

c(pairwise = sat_n_vars(pairwise), sequential = sat_n_vars(sequential))
```

The counter needs variables of its own, and models differing only in those
are the same answer repeated. Enumerated over everything, they are counted
separately; projected onto the six literals we care about, they are not:

```{r}
c(
  everything = sat_n_solutions(sat_solutions(sequential, limit = 500)),
  projected = sat_n_solutions(sat_solutions(
    sat_solver() |> (\(s) {sat_at_most(s, 1:6, 2, encoding = "sequential"); s})(),
    vars = 1:6, limit = 500
  ))
)
```

The projected figure is the number of ways to choose at most two of six,
which is `choose(6,0) + choose(6,1) + choose(6,2)`. Get into the habit of
passing `vars`: the moment an encoding grows auxiliaries, an unprojected
count stops answering your question and nothing warns you.

`sat_complete()` is worth reading every time. An enumeration stopped at
`limit` looks exactly like an exhaustive one, and the difference between
"there are 120 answers" and "there are at least 120 answers" is usually the
thing you wanted to know.

## Proving there is no answer

The Petersen graph needs three colours. Ask for two and the solver says so:

```{r}
s3 <- colouring_solver(2, edges, 10)
sat_status(sat_solve(s3))
```

`"unsat"` is a strong claim — not "I did not find one" but "none exists" —
and by default you are taking the solver's word for it. A proof turns that
into something an independent tool can check, trusting neither zusat nor
CaDiCaL:

For the proof to be checkable, the file a checker reads must contain *every*
clause the proof relies on. That rules out adding the exactly-one constraints
through `sat_exactly()` here: its clauses go into the solver, not into any
list we hold. Written out by hand they need no auxiliary variables at all —
at two colours, exactly-one is one "at least one" clause and one "not both"
clause per vertex — so the same list can go to both:

```{r}
colouring_clauses <- function(n_colours, edges, n_vertices) {
  vc <- function(v, c) (v - 1L) * n_colours + c
  clauses <- list()

  for (v in seq_len(n_vertices)) {
    lits <- vapply(seq_len(n_colours), function(c) vc(v, c), numeric(1))
    clauses <- c(clauses, list(lits))                        # at least one
    for (pair in utils::combn(n_colours, 2, simplify = FALSE)) {
      clauses <- c(clauses, list(c(-lits[pair[1]], -lits[pair[2]])))  # not both
    }
  }

  for (i in seq_len(nrow(edges))) {                          # adjacent differ
    for (c in seq_len(n_colours)) {
      clauses <- c(clauses, list(c(-vc(edges[i, 1], c), -vc(edges[i, 2], c))))
    }
  }
  clauses
}

two_colours <- colouring_clauses(2, edges, 10)
length(two_colours)
```

Now the solver and the file see exactly the same formula:

```{r}
proof <- tempfile(fileext = ".drat")
cnf <- tempfile(fileext = ".cnf")

s4 <- sat_solver()            # empty: tracing records the whole derivation
sat_trace_proof(s4, proof)    # so it must start before any clause

sat_add(s4, two_colours)
sat_status(sat_solve(s4))
sat_close_proof(s4)           # the proof is incomplete until this returns

write_dimacs(two_colours, cnf)
c(clauses = length(readLines(cnf)) - 1L, proof_steps = length(readLines(proof)))
```

Hand those two files to any DRAT checker:

```sh
drat-trim problem.cnf problem.drat
#> c 15 of 50 clauses in core
#> c 2 of 22 lemmas in core using 20 resolution steps
#> s VERIFIED
```

That is the whole point of the section. The claim "no two-colouring exists"
is now checkable by a program that has never heard of zusat or CaDiCaL.

The trade is visible in the code above: `sat_exactly()` is more convenient and
picks a better encoding as the numbers grow, but its clauses live inside the
solver. When you need a self-contained CNF to hand to another tool, build the
list yourself.

A satisfiable answer needs no proof, because the model *is* the evidence —
you check it yourself, exactly as we did above.

## Finding the chromatic number

Put those pieces together and the chromatic number falls out of a loop:

```{r}
chromatic_number <- function(edges, n_vertices, max_colours = 6) {
  for (k in seq_len(max_colours)) {
    s <- colouring_solver(k, edges, n_vertices)
    if (sat_is_sat(sat_solve(s))) {
      return(k)
    }
  }
  NA_integer_
}

chromatic_number(edges, 10)
```

Each iteration builds a fresh solver, which is the right call here because
the formula changes shape with `k`. When a search adds constraints to a
*fixed* formula — tightening a bound, fixing a decision — reuse one solver
and let it keep what it has learned.

## Guarding against a runaway solve

Colouring gets hard quickly, and a solver given a hard instance does not
return. Anywhere a hang is unacceptable, bound the search:

```{r}
s5 <- colouring_solver(3, edges, 10)
sat_limit(s5, "conflicts", 10000)

sol5 <- sat_solve(s5)
sat_status(sol5)
```

If that returns `"unknown"`, the solver gave up inside its budget. It is not
a weaker `"unsat"` and must not be read as one — the question is still open.

## What to take from this

The solver is the easy part. The parts worth care are the ones where a
mistake produces a confident wrong answer rather than an error:

- an index function that collides encodes a different problem;
- a decoded answer that nobody checks against the original constraints;
- an enumeration truncated at `limit` and read as exhaustive;
- `"unknown"` read as `"unsat"`.

Each has a cheap guard, and this article uses all four.
